---
title: "Linear_model_with_contrast"
author: "Xuening"
date: "4/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and EDA

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider.csv"
if (!file.exists(filename)) download.file(url, filename)
spider <- read.csv(filename, skip=1)
table(spider$leg,spider$type)
boxplot(spider$friction~spider$type*spider$leg,las=2,xlab="",ylab="friction") # las=2 give vertical x labels
```

## Fit a linear model
Design a model matrix that records the condition of both push/pull and leg number.
```{r}
X <- model.matrix(~type+leg+type:leg,data=spider)
head(X)
colSums(X)
```
The column sum corresponds with the table. Now fit a linear model.
```{r}
fit <- lm(friction~type*leg,data=spider)
summary(fit)
coefs <- coef(fit)
coefs
```

## Do L2 and L3 have same friction when pulling?

The average friction of L2 pulling is `coefs["(Intercept)"]`+`coefs["legL2"]`. The average friction of L3 pulling is `coefs["(Intercept)"]`+`coefs["legL3"]`. Let's verify this:
```{r message=FALSE, warning=FALSE}
coefs["(Intercept)"]+coefs["legL2"]
coefs["(Intercept)"]+coefs["legL3"]
library(dplyr)
spider %>% group_by(type, leg) %>% summarize(mean_val=mean(friction))
```

So the question equals to: is (`coefs["(Intercept)"]`+`coefs["legL2"]`)-(`coefs["(Intercept)"]`+`coefs["legL3"]`) significantly different from $0$? Since `coefs["(Intercept)"]` cancel out, we are asking whether `coefs["legL2"]`-`coefs["legL3"]` significantly different from $0$.  
  
Here contrast matrix come into play. We can write `coefs` as a column vector $\hat\beta$. Design contrast matrix as $c=[0,0,1,-1,0,0,0,0]^T$, and we find 
$$c\hat\beta=\hat\beta_3-\hat\beta_4$$
Other good properties:
$$ E[\hat\beta_3-\hat\beta_4]=E[c^T\hat\beta]=c^TE[\hat\beta]=c^T\beta=\beta_3-\beta_4$$
$$Var[\hat\beta_3-\hat\beta_4]=Var[c^T\hat\beta]=c^TVar(\hat\beta)c=\sigma^2c^T(X^TX)^{-1}c$$

Let's calculate.
```{r}
coefs["legL2"]-coefs["legL3"]

# calculate standard error
# 1. estimate variance of beta
var_beta <- sum(fit$residuals^2)/(nrow(X)-ncol(X)) * solve(crossprod(X))
head(var_beta)
```

One thing to clarify: `var_beta` is the variance-covariance matrix of all the estimated parameters.

```{r}
c <- matrix(c(0,0,1,-1,0,0,0,0))
se <- sqrt(t(c) %*% var_beta %*% c)
se
```
Then t value:
```{r}
tval <- (coefs["legL2"]-coefs["legL3"])/se
tval
```

Then p value:
```{r}
2*pt(-abs(tval),nrow(X)-ncol(X))
```

If $\alpha=0.05$ then we can conclude that L2 and L3 have significant difference in friction when pulling.  
  
We can also do the analysis above using lackage `contrast`.

```{r message=FALSE, warning=FALSE}
library(contrast)
L2pull_L3pull <- contrast(fit,
                          list(type="pull",leg="L2"),
                          list(type="pull",leg="L3"))
L2pull_L3pull
```


## Does push vs. pull have same effect on L2?

```{r}
L2push_L2pull <- contrast(fit,
                          list(type="push",leg="L2"),
                          list(type="pull",leg="L2"))
L2push_L2pull
```

So push and pull have significantly different effects on L2.  
  
This question equals to: does (`coefs["(Intercept)"]`+`coefs["legL2"]`+`coefs["typepush"]`+`coefs["typepush:legL2"]`) significantly different from (`coefs["(Intercept)"]`+`coefs["legL2"]`). After cancelling out, only `coefs["typepush"]`+`coefs["typepush:legL2"]` is left. Let's try.

```{r}
coefs["typepush"]+coefs["typepush:legL2"]
```

## Does push vs. pull have equal effects on L1 and L2?

Push-pull difference on L1 is (`coefs["(Intercept)"]`+`coefs["typepush"]`)-`coefs["(Intercept)"]`=`coefs["typepush"]`.  

Push-pull difference on L2 is (`coefs["(Intercept)"]`+`coefs["typepush"]`+`coefs["legL2"]`+`coefs["typepush:legL2"]`)-(`coefs["(Intercept)"]`+`coefs["legL2"]`)=`coefs["typepush"]`+`coefs["typepush:legL2"]`.  
  
The difference between difference is `coefs["typepush:legL2"]`. So let's see its t value and p value.

```{r}
summary(fit)$coefficients
```

The p value is 0.2144, so push vs. pull effects on L1 and L2 have no significant difference.

## Does push vs. pull have equal effects on L2 and L3?

Push-pull difference on L2 is (`coefs["(Intercept)"]`+`coefs["typepush"]`+`coefs["legL2"]`+`coefs["typepush:legL2"]`)-(`coefs["(Intercept)"]`+`coefs["legL2"]`)=`coefs["typepush"]`+`coefs["typepush:legL2"]`.   

Push-pull difference on L3 is (`coefs["(Intercept)"]`+`coefs["typepush"]`+`coefs["legL3"]`+`coefs["typepush:legL3"]`)-(`coefs["(Intercept)"]`+`coefs["legL3"]`)=`coefs["typepush"]`+`coefs["typepush:legL3"]`.  

The difference between the difference is `coefs["typepush:legL3"]`-`coefs["typepush:legL2"]`

Let validate this.

```{r message=FALSE, warning=FALSE}
library(multcomp)
C <- matrix(c(0,0,0,0,0,-1,1,0), 1)
L3vsL2interaction <- glht(fit, linfct=C)
summary(L3vsL2interaction)
```

```{r}
coefs["typepush:legL3"]-coefs["typepush:legL2"]
```

And the answer: push vs. pull have significantly different effects on L2 and L3.

## Model matrix without intercept

```{r}
spider$group <- factor(paste0(spider$leg, spider$type))
X <- model.matrix(~ 0 + group, data=spider)
head(X)
fitG <- lm(friction ~ 0 + group, data=spider)
summary(fitG)
coefs <- coef(fitG)
coefs
```

This time there is no intercepts. Thus the push vs. pull difference across L2 and L3 is just (`coefs["groupL3push"]`-`coefs["groupL3pull"]`)-(`coefs["groupL2push"]`-`coefs["groupL2pull"]`).

```{r}
C <- matrix(c(0,0,1,-1,-1,1,0,0), 1)
groupL3vsL2interaction <- glht(fitG, linfct=C)
summary(groupL3vsL2interaction)
```
This is the same as
```{r}
coefs["groupL3push"]-coefs["groupL3pull"]-(coefs["groupL2push"]- coefs["groupL2pull"])
```

