---
title: "Linear_model_with_interactions"
author: "Xuening"
date: "4/16/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load data and EDA

```{r}
url <- "https://raw.githubusercontent.com/genomicsclass/dagdata/master/inst/extdata/spider_wolff_gorb_2013.csv"
filename <- "spider.csv"
if (!file.exists(filename)) download.file(url, filename)
spider <- read.csv(filename, skip=1)
table(spider$leg,spider$type)
boxplot(spider$friction~spider$type*spider$leg,las=2,cex=0.3,xlab="",ylab="friction") # las=2 give vertical x labels
```

## Fit a linear model
Design a model matrix that records the condition of both push/pull and leg number.
```{r}
X <- model.matrix(~type+leg,data=spider)
head(X)
colSums(X)
```
The column sum corresponds with the table. Now fit a linear model.
```{r}
fit <- lm(spider$friction~X+0)
summary(fit)
coefs <- coef(fit)
coefs
```

## Interpret some coefficients

The second coefficient is the difference in mean friction between type pull and push.
```{r message=FALSE, warning=FALSE}
library(dplyr)
means <- spider %>% group_by(type) %>% summarize(mean=mean(friction))
means
```
```{r}
means$mean[2]-means$mean[1]
coefs["Xtypepush"]
```

Yet the problem lies in the first coefficient. Theoretically it should be the average friction of leg L1 and type pull because this case is represented $[1,0,0,0,0]$ in model matrix and only first coefficient can fit it.

```{r}
spider %>% group_by(type,leg) %>% summarize(mean=mean(friction))
coefs[1]
```

Oh? The mean of pull+L1 is 0.9214, not first coefficient 1.05. What happened?

## Linear model with interaction
What happened is that we assumed additive effects of `leg` and `type` thus didn't take into consideration the interactions between them. To fit a linear model with interaction, we need to redesign the model matrix.

```{r}
X <- model.matrix(~ type + leg + type:leg, data=spider)
head(X)
colSums(X)
```

`typepush:legL2` is just the product of value of `typepush` and `legL2`. Only 15 rows have value 1 for both `typepush` and `legL2`. Let's fit and see.

```{r}
fit <- lm(spider$friction~X+0)
summary(fit)
```

## Interpret all coefficients

Now let's interpret all coefficients.

```{r}
means_df <- spider %>% group_by(type,leg) %>% summarize(mean=mean(friction))
means_df
coefs <- coef(fit)
coefs
```
`coefs['X(Intercept)']` is the average of L1 pull. `coefs['Xtypepush']` is the difference between L1 push and L1 pull.
```{r}
means_df[means_df$leg=="L1" & means_df$type=="push",]$mean-means_df[means_df$leg=="L1" & means_df$type=="pull",]$mean
coefs['Xtypepush']
```
The average of L2 pull is `coefs['X(Intercept)']`+`coefs["XlegL2"]`:
```{r}
coefs['X(Intercept)']+coefs["XlegL2"]
means_df[means_df$leg=="L2" & means_df$type=="pull",]$mean
```
 The average of L2 push is `coefs['X(Intercept)']`+`coefs["XlegL2"]`+`coefs["Xtypepush:legL2"]`+ `coefs["Xtypepush"]`.
 
```{r}
coefs['X(Intercept)']+coefs["XlegL2"]+coefs["Xtypepush:legL2"]+coefs["Xtypepush"]
means_df[means_df$leg=="L2" & means_df$type=="push",]$mean
```

The average of L3 pull is `coefs['X(Intercept)']`+`coefs["XlegL3"]`:
```{r}
coefs['X(Intercept)']+coefs["XlegL3"]
means_df[means_df$leg=="L3" & means_df$type=="pull",]$mean
```
 The average of L3 push is `coefs['X(Intercept)']`+`coefs["XlegL3"]`+`coefs["Xtypepush:legL3"]`+ `coefs["Xtypepush"]`.
 
```{r}
coefs['X(Intercept)']+coefs["XlegL3"]+coefs["Xtypepush:legL3"]+coefs["Xtypepush"]
means_df[means_df$leg=="L3" & means_df$type=="push",]$mean
```

The average of L4 pull is `coefs['X(Intercept)']`+`coefs["XlegL4"]`:
```{r}
coefs['X(Intercept)']+coefs["XlegL4"]
means_df[means_df$leg=="L4" & means_df$type=="pull",]$mean
```
 The average of L4 push is `coefs['X(Intercept)']`+`coefs["XlegL4"]`+`coefs["Xtypepush:legL4"]`+ `coefs["Xtypepush"]`.
 
```{r}
coefs['X(Intercept)']+coefs["XlegL4"]+coefs["Xtypepush:legL4"]+coefs["Xtypepush"]
means_df[means_df$leg=="L4" & means_df$type=="push",]$mean
```
